---
title: "Making Figures Count"
subtitle: "A guide to the figures and R code"
author: "Aron Hill"
date: "2026-09-18"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    code-fold: true
    code-summary: "Show R code"
    code-copy: true
    code-overflow: wrap
    code-tools: false
    embed-resources: true
    theme: default
    css: pipeline.css
    fig-cap-location: bottom
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "ragg_png"
    dpi: 320
---

## Start here!

This guide follows the figures in the talk. Read the short explanations, then
open **Show R code** when you want to see how a plot was made. The same
80 simulated observations are used throughout. The guide also keeps a few
optional examples beyond the final slide sequence.

The latest slide additions are the [density plot](#a-density-plot),
[compact violin](#a-compact-violin) and [refined raincloud](#refining-the-raincloud).
For the complete talk sequence, run [J_slide_walkthrough.R](J_slide_walkthrough.R)
from `code/`. A [slide-to-code index](../presentation/Slide_code_index.md)
identifies the script and plot object for each slide.

To run a script, keep the supplied folders together. In RStudio, open the
script in `code/`, then choose **Session > Set Working Directory > To Source
File Location**. Run its lines in order. The starter needs only ggplot2.


<details class="optional">
<summary>Installation and the full pipeline</summary>

The complete pipeline uses R 4.1 or later and these packages. Install them once:

```{r}
#| label: install-packages
#| code-summary: "Package installation"
#| eval: false
install.packages(c("ggplot2", "dplyr", "readr", "patchwork", "ragg", "systemfonts"))
```

To rebuild this guide, also install Quarto and the R package `knitr`, open
[B_Figure_Pipeline.qmd](B_Figure_Pipeline.qmd), and click **Render** in RStudio.
You can read the supplied HTML without installing R or Quarto.


</details>

### A short example

`ggplot()` selects the data. `aes()` maps columns to the axes. The `geom_`
functions draw the marks, and `+` adds layers or settings. Here is the complete
starter script, including its data import and export.


```{r}
#| label: starter
#| code-summary: "Show the complete starter script"
#| eval: false

# =============================================================================
# 1. Load ggplot2 and read the data
# =============================================================================

# Install ggplot2 once if needed: install.packages("ggplot2")
library(ggplot2)
# Run from code/ so this path finds the supplied CSV.
dat <- read.csv("../data/caffeine_attention.csv")
# Set the left-to-right order of the groups.
dat$group <- factor(dat$group, levels = c("Placebo", "Caffeine 200 mg"))

# =============================================================================
# 2. Draw the participant scores
# =============================================================================

# alpha sets transparency: 0 is invisible and 1 is fully opaque.
# Try changing alpha or the jitter width, then compare with this version.
# Put group on the x-axis and each participant’s score on the y-axis.
p <- ggplot(dat, aes(x = group, y = vigilance)) +
# Move points sideways only. The seed keeps their positions repeatable.
  geom_point(position = position_jitter(width = 0.10, height = 0, seed = 1),
             alpha = 0.7) +
  labs(x = NULL, y = "Vigilance score (0-100)") +
  theme_minimal(base_size = 12)

# =============================================================================
# 3. View and save the plot
# =============================================================================

print(p)
dir.create("../figures", showWarnings = FALSE)
# Save at a specified physical size, then check that the labels are readable.
ggsave("../figures/starter_points.png", plot = p,
       width = 160, height = 115, units = "mm", dpi = 320, bg = "white")
```


For reference, `geom_col()` draws supplied bar heights, while `geom_bar()`
counts rows by default. The [ggplot2 book](https://ggplot2-book.org/getting-started)
explains the basics.

## Data and setup

This example runs through a fictional study investigating the effects of caffeine 
consumption on vigilance. The study has n = 80 participants 
(40 participants per group: placebo or 200 mg caffeine).
Vigilance scores range from 0 to 100, with higher scores indicating better
performance. The CSV contains `id`, `group`, `metaboliser` and `vigilance`.
The metaboliser labels and their relationship to scores are invented.

The full pipeline uses two tables: **dat** has all 80 participant rows;
**summ** has one summary row per group. Run the setup before the figure chunks.


```{r}
#| label: sync-downloads
#| code-summary: "Show R code"
#| include: false
source("C_sync_scripts.R", local = new.env())
```

```{r}
#| label: setup
#| code-summary: "Show data import and shared settings"
# Install these once if needed:
# install.packages(c("ggplot2", "dplyr", "readr", "patchwork", "ragg", "systemfonts"))

# Packages used throughout the guide.
library(ggplot2)
library(dplyr)
library(readr)
library(patchwork)


# Use an installed font so the code also runs on another computer.
pick_font <- function(candidates = c("Lato", "Source Sans Pro", "Helvetica Neue",
                                     "Helvetica", "Arial", "sans")) {
  installed <- systemfonts::system_fonts()$family
  candidates[candidates %in% c(installed, "sans")][1]
}
ft <- pick_font()

# Read one row per participant and keep the treatment groups in this order.
dat <- read_csv("../data/caffeine_attention.csv", show_col_types = FALSE) |>
  mutate(group = factor(group, levels = c("Placebo", "Caffeine 200 mg")),
         xnum  = as.numeric(group))

# group_by calculates the summaries separately for each treatment.
# dat keeps the participant scores; summ has only one row per group.
# se is the standard error; ci is the half-width of a 95% t interval.
summ <- dat |>
  group_by(group) |>
  summarise(n = n(), mean = mean(vigilance), sd = sd(vigilance),
            se = sd / sqrt(n), ci = qt(0.975, n - 1) * se, .groups = "drop") |>
  mutate(xnum = as.numeric(group))

# These summaries are used in the optional results table.
descr <- dat |>
  group_by(group) |>
  summarise(n = n(), median = median(vigilance),
            q1 = quantile(vigilance, 0.25), q3 = quantile(vigilance, 0.75),
            .groups = "drop")

tt <- t.test(vigilance ~ group, data = dat)

# Reuse the same group colours in every figure.
ok <- c(Placebo = "#4B77A8", `Caffeine 200 mg` = "#D98C1F")
ink <- "grey20"; ink2 <- "grey38"; rule <- "grey88"

# Themes control appearance without changing the observations or summaries.
# base_size sets a starting text size; the offsets below adjust each label.
# Shared formatting. Change it here to apply it across the figure sequence.
make_theme_talk <- function(base_size = 11) {
  theme_minimal(base_size = base_size, base_family = ft) +
    theme(
      plot.title          = element_text(size = base_size + 1.5, face = "bold",
                                         colour = ink, margin = margin(b = 3)),
      plot.subtitle       = element_text(size = base_size - 1.2, colour = ink2,
                                         margin = margin(b = 14)),
      plot.caption        = element_text(size = base_size - 2.7, colour = ink2,
                                         hjust = 0, margin = margin(t = 12),
                                         lineheight = 1.25),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      axis.title          = element_text(size = base_size - 1.2, colour = ink2),
      axis.title.y        = element_text(margin = margin(r = 9)),
      axis.title.x        = element_text(margin = margin(t = 9)),
      axis.text           = element_text(size = base_size - 1.7, colour = ink2),
      axis.ticks          = element_blank(),
      panel.grid.minor    = element_blank(),
      panel.grid.major.x  = element_blank(),
      panel.grid.major.y  = element_line(colour = rule, linewidth = 0.3),
      legend.text         = element_text(size = base_size - 1.7, colour = ink2),
      legend.key.height   = unit(11, "pt"),
      plot.margin         = margin(13, 15, 11, 13)
    )
}
theme_talk    <- make_theme_talk(11)   # figures 1-5
theme_talk_lg <- make_theme_talk(13)   # figures 6 and 6b
theme_talk_xl <- make_theme_talk(14) + # figures 7 and 8 - publication-ready
  theme(panel.border = element_rect(colour = "grey55", fill = NA, linewidth = 0.6))

# Width and height passed to this helper are in inches. ragg gives consistent
# text and line rendering for the PNGs. Check the result at its intended size.
# Save a PNG in figures/ using the same graphics device as the guide.
save_fig <- function(name, plot, w, h) {
  dir.create("../figures", showWarnings = FALSE)
  ggsave(file.path("../figures", name), plot, width = w, height = h,
         dpi = 320, bg = "white", device = ragg::agg_png)
  invisible(plot)
}

YL <- "Vigilance score (0-100)"

# This score range is specific to the example dataset.
score_axis <- function() scale_y_continuous(limits = c(35, 85), breaks = seq(40, 80, 10))
```


## 1. Bar charts

### The first draft

The bars show means, but the image leaves the error bars undefined and hides
individual scores. Here the intervals are ±1 standard error (SEM), which
describes precision of the mean, not the spread of participants.


```{r}
#| label: fig-badbar
#| code-summary: "Show R code"
#| fig-width: 4.6
#| fig-height: 3.8
#| fig-cap: |
#|   Group means with error bars of ±1 SEM. Synthetic data, n = 40 per group. The chart does not identify its error bars or sample size on the image itself. Its small labels and raw variable names make it harder to read.

# The table summ contains the group means and standard errors.
p1 <- ggplot(summ, aes(x = group, y = mean)) +
  # Draw one bar at each supplied mean.
  geom_col(fill = "grey60", width = 0.6) +
  # Error bars extend one standard error either side of the mean.
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.15) +
  labs(x = "group", y = "vigilance") +
  theme_grey(base_size = 11, base_family = ft)

# Small labels and a full 0–100 axis recreate the first draft.
p1_bad <- p1 +
  scale_y_continuous(limits = c(0, 100)) +
  theme_grey(base_size = 7, base_family = ft)

p1_bad
save_fig("01_bad_bar.png", p1_bad, 4.6, 3.8)
```

<details class="optional">
<summary>Axis limits: the same bars with different baselines</summary>

Starting the axis at 55 exaggerates the difference in visible bar lengths.
The values themselves do not change. Bars need a zero baseline when length
represents magnitude. Position-based plots, such as points and boxes, can use
a suitable restricted range.

```{r}
#| label: fig-truncated
#| code-summary: "Show R code"
#| fig-width: 8.4
#| fig-height: 3.8
#| fig-cap: |
#|   The same group means and ±1 SEM error bars with two axis ranges. The right panel truncates the score axis at 55, exaggerating the difference in visible bar lengths. The numerical means are unchanged. Synthetic data, n = 40 per group.
# Run the first-bar section first: p1 is reused below.
# coord_cartesian zooms the display without removing data or recomputing means.
lab <- theme(plot.title = element_text(size = 10.5, family = ft, colour = ink))

# Reuse the same plot; only the visible axis range changes.
p2 <- (p1 + labs(title = "y axis starts at 0") + lab) +
  (p1 + coord_cartesian(ylim = c(55, 65)) +
       labs(title = "y axis starts at 55") + lab)

p2
save_fig("02_truncated_axis.png", p2, 8.4, 3.8)
```

</details>


### Labels and confidence intervals

This version identifies the outcome and sample size. It also changes the
intervals to **95% confidence intervals**, so the statistical information has
changed too. In the code, `ci` is the interval half-width calculated in `summ`.


```{r}
#| label: fig-honestbar
#| code-summary: "Show R code"
#| fig-width: 5.2
#| fig-height: 4.3
#| fig-cap: |
#|   Group means and 95% confidence intervals, calculated separately for each group using a t distribution. Synthetic data, n = 40 per group. The chart describes the means and their uncertainty but does not show the distribution of participant scores.
# Map fill to group so the two bars have consistent colours.
p3 <- ggplot(summ, aes(x = group, y = mean, fill = group)) +
  geom_col(width = 0.5) +
  # Use the 95% CI half-width instead of the standard error.
  geom_errorbar(aes(ymin = mean - ci, ymax = mean + ci), width = 0.1,
                linewidth = 0.5, colour = ink) +
  # The x-axis already names the groups, so no legend is needed.
  scale_fill_manual(values = ok, guide = "none") +
  scale_y_continuous(limits = c(0, 90), breaks = seq(0, 90, 15),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = YL,
       title = "Group means and uncertainty",
       subtitle = "Group means with 95% confidence intervals, n = 40 per group") +
  theme_talk

p3
save_fig("03_honest_bar.png", p3, 5.2, 4.3)
```


The chart still tells us little about the distribution of individual scores.

## 2. Boxplots

A boxplot uses all participant rows in `dat`. The line is the median and the
box spans the middle 50% of scores. Whiskers end at observations within
1.5 interquartile ranges of each quartile. Values beyond them are not
automatically errors.


```{r}
#| label: fig-boxplot
#| code-summary: "Show R code"
#| fig-width: 5.2
#| fig-height: 4.3
#| fig-cap: |
#|   Median (line), interquartile range (box), and whiskers extending to the most extreme observations within 1.5 × IQR of each quartile. Points beyond the whiskers indicate potential outliers. Synthetic data, n = 40 per group.
# alpha controls the fill transparency; width changes how wide the boxes look.
# Neither setting changes the median, quartiles or whiskers. A separate symbol
# marks scores beyond the whiskers; that rule does not establish a data error.
# Use participant-level data, rather than the summary table.
p4 <- ggplot(dat, aes(x = group, y = vigilance, fill = group)) +
  geom_boxplot(width = 0.32, alpha = 0.5, linewidth = 0.4, colour = ink,
               outlier.shape = 21, outlier.size = 1.6, outlier.alpha = 0.7) +
  scale_fill_manual(values = ok, guide = "none") +
  score_axis() +
  labs(x = NULL, y = YL,
       title = "A box plot",
       subtitle = "Five numbers instead of one: median, quartiles, whiskers") +
  theme_talk

p4
save_fig("04_boxplot.png", p4, 5.2, 4.3)
```


The caffeine group has a wider spread. A boxplot compresses the distribution,
so we need another view to see its shape. It does not assume normality.


<details class="optional">
<summary>Summary statistics and the mean comparison</summary>

```{r}
#| label: tbl-summary
#| code-summary: "Show R code"
#| tbl-cap: "Group summary statistics."
bind_cols(
  descr,
  summ |> select(mean, sd)
) |>
  transmute(Group = group, N = n, Mean = round(mean, 1), SD = round(sd, 1),
            Median = round(median, 1),
            IQR = paste0(round(q1, 1), "–", round(q3, 1))) |>
  knitr::kable()
```

```{r}
#| label: ttest
#| code-summary: "Show R code"
tt
```

The group SDs describe participant spread. The t-test addresses the means.
A non-significant result does not establish equivalence. 


</details>


## 3. Histograms and violins

The histogram counts observations in 4-point bins. The two panels use the
same scales. This guide also shows a smooth density curve and the group mean;
the slide version omits those overlays.


```{r}
#| label: fig-histogram
#| code-summary: "Show R code"
#| fig-width: 5.4
#| fig-height: 4.8
#| fig-cap: |
#|   Histograms with 4-point bins, a boundary at zero, and a common count scale. Curves show kernel density estimates rescaled to counts; dashed lines mark group means. Synthetic data, n = 40 per group. The caffeine distribution has two visible concentrations of scores.
# Try a bin width of 2, 4 or 8 and compare the shapes. Recalculate dens5 as
# well: the density overlay is multiplied by sample size and bin width so it
# can be displayed on the histogram count scale.
binwidth5 <- 4

dens5 <- dat |>
  group_by(group) |>
  group_modify(~ {
    d <- density(.x$vigilance, adjust = 0.9)
    tibble(x = d$x, y = d$y * nrow(.x) * binwidth5)
  }) |>
  ungroup()

p5 <- ggplot(dat, aes(x = vigilance)) +
  geom_histogram(aes(fill = group), binwidth = binwidth5, boundary = 0, colour = "white",
                 linewidth = 0.4, alpha = 0.55) +
  geom_line(data = dens5, aes(x = x, y = y, colour = group), linewidth = 0.9) +
  geom_vline(data = summ, aes(xintercept = mean), linetype = "22",
             colour = ink, linewidth = 0.4) +
  # Use a separate panel for each treatment with the same axes.
  facet_wrap(~ group, ncol = 1, scales = "fixed") +
  scale_fill_manual(values = ok, guide = "none") +
  scale_colour_manual(values = ok, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = YL, y = "Participants",
       title = "A histogram",
       subtitle = "Bin width 4 points, curve: kernel density (rescaled).\nDashed line marks the group mean.") +
  theme_talk +
  theme(panel.grid.major.x = element_line(colour = rule, linewidth = 0.3),
        strip.text = element_text(face = "bold", hjust = 0, size = 9.8,
                                  colour = ink, margin = margin(b = 4)),
        panel.spacing.y = unit(14, "pt"))

p5
save_fig("05_histogram.png", p5, 5.4, 4.8)
```


Bin width and smoothing affect the appearance. Check reasonable alternatives
against the individual observations before interpreting apparent peaks.


### A density plot

A density curve gives a smooth estimate of the distribution. Its height is
**density**, not a participant count. The full curve for each group has area
one. Small marks along the bottom (a *rug*) show individual scores.

The caffeine curve has two broad peaks at this smoothing setting. Try
`adjust = 0.75` and `1.5`: the apparent peaks can merge with stronger smoothing.
Use the histogram and individual points to check the impression. This sample
alone does not establish that the population has two distinct subgroups.

```{r}
#| label: slide-additions-setup
#| include: false
# These scripts include their own data import, plot settings and exports.
# Run K or L directly to work on one of these examples in RStudio. Here the
# separate environments prevent their settings from replacing guide objects.
# Isolate the slide objects from the earlier guide examples.
slide_density <- new.env()
sys.source("K_density_plot.R", envir = slide_density)
slide_finishes <- new.env()
sys.source("L_plot_finishes.R", envir = slide_finishes)
```

```{r}
#| label: density-script
#| code-summary: "Show the complete density script"
#| eval: false
#| code: !expr readLines("K_density_plot.R")
```

```{r}
#| label: fig-slide-density
#| echo: false
#| fig-width: 6.4
#| fig-height: 5.3
#| fig-cap: "Density estimates and individual rug marks for the same synthetic observations (n = 40 per group). Both panels use the same axes; adjust = 1."
slide_density$p_density
```

Download [K_density_plot.R](K_density_plot.R), or use the supplied
[PNG](../figures/slide_walkthrough/density_plot.png) and
[PDF](../figures/slide_walkthrough/density_plot.pdf).


<details class="optional">
<summary>The violin on its own</summary>

A violin shows a smoothed density: wider sections indicate more scores
nearby. `adjust` controls smoothing. This version marks the median with a point.

```{r}
#| label: fig-violin
#| code-summary: "Show R code"
#| fig-width: 5.6
#| fig-height: 4.5
#| fig-cap: |
#|   Violin plots show kernel density estimates with bandwidth adjustment 0.8; the point marks the median. Equal-area scaling permits comparison of distribution shapes here, with n = 40 per group. Synthetic data.
# alpha controls transparency; width sets the available horizontal space.
# trim = TRUE stops the violin at the observed minimum and maximum.
# adjust controls smoothing; smaller values show more detail.
p6 <- ggplot(dat, aes(x = group, y = vigilance, fill = group)) +
  geom_violin(width = 0.62, alpha = 0.55, colour = ink, linewidth = 0.35,
              trim = TRUE, adjust = 0.8) +
  stat_summary(fun = median, geom = "point", size = 2.6, colour = "white") +
  stat_summary(fun = median, geom = "point", size = 1.5, colour = ink) +
  scale_fill_manual(values = ok, guide = "none") +
  score_axis() +
  labs(x = NULL, y = YL,
       title = "A violin plot",
       subtitle = "Width shows estimated density. Point marks the median.") +
  theme_talk_lg

p6
save_fig("06_violin.png", p6, 5.6, 4.5)
```

</details>


### Violin with a boxplot

The violin shows shape, while the narrow box adds the median and interquartile
range. These full violins use equal-area scaling, with equal group sizes.


```{r}
#| label: fig-violinbox
#| code-summary: "Show R code"
#| fig-width: 5.6
#| fig-height: 4.5
#| fig-cap: |
#|   Violin plots with embedded boxplots. The density shows shape; the box marks the median and IQR, with whiskers to observations within the 1.5 × IQR fences. Synthetic data, n = 40 per group.
# The box and violin use the same participant rows. The violin adds shape;
# the box adds median and quartiles. outlier.shape = NA hides separate boxplot
# outlier symbols here, while the violin still uses all the observations.
# Draw the violin first so the narrow box remains visible on top.
p6b <- ggplot(dat, aes(x = group, y = vigilance, fill = group)) +
  geom_violin(width = 0.62, alpha = 0.4, colour = ink, linewidth = 0.3,
              trim = TRUE, adjust = 0.8) +
  geom_boxplot(width = 0.14, fill = "white", colour = ink, linewidth = 0.4,
               outlier.shape = NA) +
  scale_fill_manual(values = ok, guide = "none") +
  score_axis() +
  labs(x = NULL, y = YL,
       title = "Violin + box",
       subtitle = "Distribution shape, median and interquartile range.") +
  theme_talk_lg

p6b
save_fig("06b_violin_box.png", p6b, 5.6, 4.5)
```


### A compact violin

For a smaller figure, replace the box with a median dot and two simple lines.
The **white dot is the median**, the **thick line is the middle 50% (IQR)**,
and the **thin line is the observed minimum–maximum**. Neither line is a
confidence interval. The thin line uses a different definition from the
1.5 × IQR whiskers in the preceding boxplot.

The smooth outline retains the distribution shape, but individual observations
are no longer visible. Prefer the points or raincloud version when individual
scores matter to the question.

```{r}
#| label: fig-compact-slide
#| echo: false
#| fig-width: 6.30
#| fig-height: 4.92
#| fig-cap: "Compact violin for synthetic vigilance scores, n = 40 per group. White dot: median. Thick line: IQR. Thin line: observed range. Density smoothing uses adjust = 0.8."
slide_finishes$p_compact
```

```{r}
#| label: compact-slide-run
#| eval: false
# Run from code/. This loads the data, calculates the summaries and exports
# both the compact violin and the refined raincloud.
source("L_plot_finishes.R")
p_compact
```

The complete code, including `violin_summary` and `violin_style`, is in
[L_plot_finishes.R](L_plot_finishes.R) and under
[Refining the raincloud](#refining-the-raincloud) below.
Download the [160 mm PNG](../figures/plot_finishes/violin_compact.png) or
[PDF](../figures/plot_finishes/violin_compact.pdf).
A separately typeset **90 × 100 mm** version is also supplied as
[PNG](../figures/plot_finishes/violin_compact_90mm.png) and
[PDF](../figures/plot_finishes/violin_compact_90mm.pdf). It keeps readable physical
text sizes instead of simply shrinking the larger image.


## 4. Individual observations

Points can overlap. Horizontal jitter separates them while `height = 0`
keeps each score unchanged. A fixed `seed` makes the arrangement repeatable.


```{r}
#| label: fig-points
#| code-summary: "Show R code"
#| fig-width: 10.5
#| fig-height: 3.8
#| fig-cap: |
#|   The same 80 observations shown without jitter, with horizontal jitter, and with a boxplot added. Vertical positions never change. Boxplot outlier symbols are suppressed only because all scores already appear in the point layer.
# The fixed jitter seed makes the point positions repeatable. height = 0
# matters: all movement should be sideways, preserving the measured scores.
# Try changing the width to see when the points become easier to distinguish.
# Share axes and labels across the three panels.
point_base <- ggplot(dat, aes(x = group, y = vigilance)) +
  scale_colour_manual(values = ok, guide = "none") +
  scale_x_discrete(labels = c("Placebo\n(n = 40)", "Caffeine 200 mg\n(n = 40)")) +
  coord_cartesian(ylim = c(35, 85)) +
  labs(x = NULL, y = YL) + theme_talk
p_points0 <- point_base + geom_point(aes(colour = group), alpha = 0.7, size = 2)
# Horizontal jitter separates overlaps without changing any scores.
p_points <- point_base +
  geom_point(aes(colour = group), alpha = 0.7, size = 2,
             position = position_jitter(width = 0.10, height = 0, seed = 1))
# Draw the box first, then the points. Hide duplicate outlier symbols.
p_points_box <- point_base +
  geom_boxplot(width = 0.22, fill = "white", colour = ink,
               outlier.shape = NA, linewidth = 0.4) +
  geom_point(aes(colour = group), alpha = 0.7, size = 2,
             position = position_jitter(width = 0.10, height = 0, seed = 1))
p_points_sequence <- (p_points0 + labs(title = "1. Overlapping points")) +
  (p_points + labs(title = "2. Horizontal jitter")) +
  (p_points_box + labs(title = "3. Points + box"))
p_points_sequence
save_fig("06c_points_sequence.png", p_points_sequence, 10.5, 3.8)
save_fig("06c_points.png", p_points, 5.2, 4.3)
save_fig("06d_points_box.png", p_points_box, 5.2, 4.3)
```


`outlier.shape = NA` hides duplicate boxplot outlier symbols because the point
layer already shows every observation. It does not remove data. This points-and-box
plot is already a useful option for a paper.

## 5. Raincloud plots

A raincloud places the observations, boxplot and density beside one another.
The clouds here are each scaled to their own maximum width: compare their
shapes, not absolute density between groups.


```{r}
#| label: fig-raincloud
#| code-summary: "Show R code"
#| fig-width: 6.2
#| fig-height: 4.8
#| fig-cap: |
#|   Vigilance scores after placebo or 200 mg caffeine in a fictional sustained-attention study (n = 40 per group). Higher scores indicate better performance. Points show individual participants with horizontal jitter only. Half-violins show kernel density estimates (bandwidth adjustment 0.8), each scaled to its own maximum width. Boxes show medians and IQRs; whiskers extend to the most extreme observations within 1.5 × IQR of each quartile. All data are synthetic.
# density() returns score positions and their estimated densities. The first
# and last zero-width points close the polygon along the group position.
# Changing adjust changes smoothing, so compare with the individual scores.
# Estimate each cloud and scale it to its own maximum width.
cloud <- dat |>
  group_by(group) |>
  group_modify(~ {
    d <- density(.x$vigilance, adjust = 0.8,
                 from = min(.x$vigilance), to = max(.x$vigilance))
    tibble(v = c(d$x[1], d$x, d$x[length(d$x)]),
           w = c(0, d$y / max(d$y), 0))
  }) |>
  ungroup() |>
  mutate(xnum = as.numeric(group))

# Dividing by max(d$y) scales each cloud to its own widest point. Compare
# distribution shapes here, rather than absolute density across groups.
# Horizontal offsets place the cloud, box and points beside one another.
rain_layers <- function(point_aes) list(
  geom_polygon(data = cloud, aes(x = xnum + 0.05 + w * 0.26, y = v, fill = group),
               colour = NA, alpha = 0.5),
  # group is explicit because the horizontal positions are numeric.
  geom_boxplot(data = dat, aes(x = xnum - 0.055, y = vigilance, group = group),
               width = 0.06, fill = "white", alpha = 0.9, outlier.shape = NA,
               linewidth = 0.4, colour = ink),
# The points sit to the left of the box and remain visible as their own layer.
  # height = 0 preserves scores; seed = 1 makes the jitter repeatable.
  geom_point(data = dat, point_aes, size = 2.3, alpha = 0.78, stroke = 0,
             position = position_jitter(width = 0.042, height = 0, seed = 1)),
  scale_x_continuous(breaks = 1:2, labels = paste0(levels(dat$group), "\n(n = 40)"),
                     limits = c(0.72, 2.36))
)

# Assemble the layers, labels and shared formatting.
p7 <- ggplot() +
  rain_layers(aes(x = xnum - 0.19, y = vigilance, colour = group)) +
  scale_colour_manual(values = ok, guide = "none") +
  scale_fill_manual(values = ok, guide = "none") +
  labs(x = NULL, y = YL,
       title = "A raincloud plot",
       subtitle = "Rain: every participant. Cloud: kernel density. Box: median and IQR.") +
  score_axis() +
  theme_talk_xl

p7
save_fig("07_raincloud.png", p7, 6.2, 4.8)
```


The full caption is also in `figures/07_raincloud_caption.txt`.
To make this figure without running the earlier examples, use
[D_final_raincloud.R](D_final_raincloud.R). It includes its own setup.


<details class="optional">
<summary>How the half-violin works</summary>

`density()` estimates the curve and `geom_polygon()` draws one side.
The offsets separate the cloud, box and points. Zero-width endpoints close
the polygon. The boxplot has an explicit `group` mapping because its x
coordinates are numeric. These details are included in the full code above.


</details>


### Refining the raincloud

Keep the same scores and density estimate, then make the figure easier to read:
increase label sizes, widen the boxes slightly, and use larger filled points
with thin white outlines. A little more horizontal jitter helps separate the
points; `height = 0` keeps every score unchanged. Light horizontal gridlines
help readers compare scores without competing with the data.

```{r}
#| label: finishes-script
#| code-summary: "Show the complete raincloud and compact violin script"
#| eval: false
#| code: !expr readLines("L_plot_finishes.R")
```

```{r}
#| label: fig-refined-slide
#| echo: false
#| fig-width: 6.30
#| fig-height: 4.92
#| fig-cap: "Refined raincloud at 160 × 125 mm. Points: all 80 synthetic observations with horizontal jitter only. Boxes: median and IQR, with whiskers to observations within the 1.5 × IQR fences. Clouds: density estimates (adjust = 0.8), each scaled to its own maximum width."
slide_finishes$p_polished
```

Run [L_plot_finishes.R](L_plot_finishes.R) from `code/`; it loads
[I_slide_setup.R](I_slide_setup.R) automatically. The script exports the
refined raincloud and both sizes of compact violin to `figures/plot_finishes/`.
Download the refined raincloud as [PNG](../figures/plot_finishes/raincloud_refined.png)
or [PDF](../figures/plot_finishes/raincloud_refined.pdf).
Check the export at its intended size and adapt it to the journal's requirements.


## 6. The simulated subgroups

The caffeine data came from two simulated populations: 22 participants with
a higher mean and 18 with a lower mean. We assigned fictional fast/slow
metaboliser labels to them.


```{r}
#| label: fig-reveal
#| code-summary: "Show R code"
#| fig-width: 6.2
#| fig-height: 5.0
#| fig-cap: |
#|   The same synthetic scores, with points coloured by the subgroup labels used to generate the caffeine data (22 labelled fast and 18 labelled slow metabolisers). Grey clouds and boxplots still summarise the original treatment groups. The labels and their relationship to performance are invented for this demonstration; this is not evidence about caffeine metabolism.
dat2 <- dat |> mutate(sub = ifelse(is.na(metaboliser), "Placebo",
                                   paste(metaboliser, "metaboliser")),
                      sub = factor(sub, levels = c("Placebo", "Fast metaboliser",
                                                   "Slow metaboliser")))
pal2 <- c("Placebo" = "#4B77A8", "Fast metaboliser" = "#2E8B6F", "Slow metaboliser" = "#B5628F")

# This example reuses the raincloud objects from the preceding section.
# The subgroup labels were invented when simulating the data. They illustrate
# how known group membership can explain a pattern, not how to infer a
# biological mechanism from the shape of a plot.
p8 <- ggplot() +
  geom_polygon(data = cloud, aes(x = xnum + 0.05 + w * 0.26, y = v, group = group),
               fill = "grey72", colour = NA, alpha = 0.4) +
  geom_boxplot(data = dat, aes(x = xnum - 0.055, y = vigilance, group = group),
               width = 0.06, fill = "white", alpha = 0.9, outlier.shape = NA,
               linewidth = 0.4, colour = ink) +
  geom_point(data = dat2, aes(x = xnum - 0.19, y = vigilance, colour = sub),
             size = 2.3, alpha = 0.8, stroke = 0,
             position = position_jitter(width = 0.042, height = 0, seed = 1)) +
  scale_colour_manual(values = pal2, name = NULL) +
  scale_x_continuous(breaks = 1:2, labels = paste0(levels(dat$group), "\n(n = 40)"), limits = c(0.72, 2.36)) +
  labs(x = NULL, y = YL,
       title = "The reveal",
       subtitle = "Invented subgroup labels used to generate these synthetic data") +
  score_axis() +
  theme_talk_xl +
  theme(legend.position = "top", legend.justification = "left",
        legend.key.height = unit(13, "pt"),
        legend.background = element_rect(fill = "white", colour = "grey70",
                                         linewidth = 0.35),
        legend.margin = margin(t = 6, b = 6, l = 9, r = 9),
        legend.box.margin = margin(b = 4))

p8
save_fig("08_reveal.png", p8, 6.2, 5.0)
```


We know the explanation because we created the data. A similar pattern in a
real study could suggest a hypothesis, but would not identify its cause.
A subgroup discovered from the same outcomes remains exploratory.

## 7. Manuscript figures

### Size and export

Check the saved figure at its intended size. Here it is **160 × 125 mm**;
use your journal's requirements. `ggsave()` names the plot and sets its
physical dimensions. More DPI cannot fix lettering that is too small.


```{r}
#| label: fig-manuscript
#| code-summary: "Show R code"
#| fig-width: 6.30
#| fig-height: 4.92
#| fig-cap: |
#|   Manuscript example at a specified export size: 160 × 125 mm. The same synthetic data and raincloud layers are retained; typography is set for this output. A greyscale version and a PDF are also supplied for checking and reuse.
# Run the raincloud section first: this starts from p7.
# A larger DPI makes a sharper PNG but does not make small labels readable.
# Adjust the text at the physical size required by the manuscript.
# Keep the data layers and set text sizes for the manuscript export.
p_paper <- p7 + labs(title = NULL, subtitle = NULL) +
  theme_minimal(base_size = 10, base_family = ft) +
  theme(axis.title = element_text(size = 10, colour = ink),
        axis.text = element_text(size = 9, colour = ink),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.25),
        plot.margin = margin(6, 6, 6, 6))
p_paper
# plot = names the exact object to save; width and height are in millimetres.
ggsave("../figures/09_manuscript.png", plot = p_paper,
       width = 160, height = 125, units = "mm", dpi = 320,
       bg = "white", device = ragg::agg_png)
ggsave("../figures/09_manuscript.pdf", plot = p_paper,
       width = 160, height = 125, units = "mm", bg = "white",
       device = grDevices::cairo_pdf)
# PDF preserves vector lines and text. PNG is convenient for placing in slides.
# Check whether group labels still make the figure clear without colour.
p_paper_grey <- p_paper +
  scale_colour_manual(values = c(Placebo = "grey30", `Caffeine 200 mg` = "grey55"),
                      guide = "none") +
  scale_fill_manual(values = c(Placebo = "grey30", `Caffeine 200 mg` = "grey55"),
                    guide = "none")
ggsave("../figures/09_manuscript_greyscale.png", plot = p_paper_grey,
       width = 160, height = 125, units = "mm", dpi = 320,
       bg = "white", device = ragg::agg_png)
```


The code saves a PNG, a vector PDF and a greyscale check. Keep group labels
as well as colour, and explain the sample size and every layer in the caption.


<details class="optional">
<summary>Alternative figure: violin, points and group means</summary>

This version uses means and 95% confidence intervals instead of medians and
interquartile ranges. Those summaries describe different things. The caffeine
mean falls between the two concentrations of scores.

```{r}
#| label: fig-alternative
#| code-summary: "Show R code"
#| fig-width: 5.6
#| fig-height: 4.6
#| fig-cap: |
#|   Alternative final figure for the same synthetic data. Violin: estimated density, trimmed to the observed range. Points: individual participants, jittered horizontally with scores unchanged. White circle and black bar: group mean and its 95% confidence interval. n = 40 per group.
# This summary uses the group means and their 95% confidence intervals from
# summ. These intervals describe mean uncertainty, not participant spread.
# The mean can sit in a sparse area between clusters of observations.
# Draw the violin, then the raw scores, then the mean and confidence interval.
alt_layers <- list(
  geom_violin(aes(fill = group), width = 0.72, alpha = 0.32, colour = NA,
              trim = TRUE, adjust = 0.9),
  geom_point(aes(colour = group), size = 2.1, alpha = 0.8, stroke = 0,
             position = position_jitter(width = 0.085, height = 0, seed = 3)),
  # This layer uses summ, not the participant-level table.
  geom_errorbar(data = summ, aes(x = group, ymin = mean - ci, ymax = mean + ci),
                inherit.aes = FALSE, width = 0, linewidth = 0.9, colour = "black"),
  # Shape 21 allows a white centre and a separate dark outline.
  geom_point(data = summ, aes(x = group, y = mean), inherit.aes = FALSE,
             shape = 21, size = 3.6, stroke = 0.9, fill = "white", colour = "black"),
  scale_fill_manual(values = ok, guide = "none"),
  scale_colour_manual(values = ok, guide = "none"),
  scale_x_discrete(labels = paste0(levels(dat$group), "\n(n = 40)")),
  score_axis(),
  labs(x = NULL, y = YL)
)

p_alt <- ggplot(dat, aes(x = group, y = vigilance)) + alt_layers +
  labs(title = "Violin with raw data and group mean",
       subtitle = "White circle: group mean. Black bar: 95% confidence interval.") +
  theme_talk_xl
p_alt
save_fig("12_violin_points_ci.png", p_alt, 5.6, 4.6)
```

The next version changes only the theme, with drawn axes and no gridlines.
Choose the style that suits the publication.

```{r}
#| label: fig-alternative-journal
#| code-summary: "Show R code"
#| fig-width: 5.6
#| fig-height: 4.6
#| fig-cap: |
#|   The same figure with drawn axes, outward tick marks and no gridlines. Only the theme differs; every data layer is identical to the previous figure.
# Run the preceding alternative-figure section first to create alt_layers.
# Use this comparison to judge appearance while keeping the data layers fixed.
theme_journal <- theme_talk_xl +
  theme(panel.grid.major.y = element_blank(),
        axis.line  = element_line(colour = ink, linewidth = 0.4),
        axis.ticks = element_line(colour = ink, linewidth = 0.4),
        axis.ticks.length = unit(3.5, "pt"))

# Reuse every data layer and change only the theme.
p_alt_journal <- ggplot(dat, aes(x = group, y = vigilance)) + alt_layers +
  labs(title = "Violin with raw data and group mean",
       subtitle = "Journal theme: drawn axes, no gridlines.") +
  theme_journal
p_alt_journal
save_fig("12b_violin_points_ci_journal.png", p_alt_journal, 5.6, 4.6)
```

</details>


## Using your own data

Replace the CSV import and the columns mapped in `aes()`. Check missing values,
group sizes and units. The fixed `n = 40` labels and 35–85 score range belong
to this example. Choose the summary and interval for your question, then inspect
the export at its final size.
